---
title: "Data analysis and visualization using R"
author: "Michiel Noback"
date: "november 2015"
output:
  ioslides_presentation: default
  beamer_presentation: default
  pdf_document: default
subtitle: Complex data types
job: lecturer at Hanze University of Applied Science
---

```{r setup, cache = F, echo = F, message = F, warning = F}
#comment this out for pdf output
source("rmarkdown_knitr_header.R")
```

```{r echo = F}
par(mar=c(3,3,0,0))
```


## Complex datatypes and IO

**Contents**

- Matrices
- Factors
- Lists
- Data frames
- Reading dataframes from file (first iteration)
- Plotting with dataframes

# Matrices

## Matrices are vectors with dimensions

- We will not detail on them in this course, only this one slide
- This does not mean they are not important, but they are just not the focus here


```{r matrix_demo}
m <- matrix(1:10, nrow = 2, ncol = 5); m
v <- 1:10; dim(v) <- c(2, 5); v
```


# Factors

## Factors

- Although factors are not really complex, I saved them because they have some strange behaviour.
- Factors are used to represent different levels of some explanatory variable. 
- For instance:  
    - eye color (brown, blue, green)
    - weight class (underweight, normal, obese)
    - plant age in years (1, 2, 3)


## Factors
- Factors are used to represent data in nominal or ordinal scales
- Nominal has no order; Ordinal has
- these functions are relevant 
    - `factor(x)` 
    - `as.factor(x)`
    - `factor(x, levels = my_levels, labels = my_labels)`

## Character to factor

- Suppose you have surveyed the eye color of your class room and found these values

```{r eye_color_1, tidy=FALSE}
eye_colors <- c("green", "blue", "brown", "brown", "blue",
    "brown", "brown", "brown", "blue", "brown", "green",
    "brown", "brown", "blue", "blue", "brown")
```

- next you would like to plot or tabulate these findings


## Plot character data

- Simply plotting gives an error

```{r eye_color_2, error=TRUE}
plot(eye_colors)
```


## Plot factor data

- Plotting a character vector converted to a factor is easy

```{r eye_color_3, fig.width=4.5, fig.height=3.5}
eye_colors <- as.factor(eye_colors)
plot(eye_colors)
```


## Tabulate factor data

- Factors are also really easy to tabulate and filter

```{r eye_color_4}
table(eye_colors)
sum(eye_colors == "blue")
```


## Defining levels

- Especially when working with ordinal scales, defining the order of the factors (levels) is useful
- By default, R uses the natural ordering (numerical/alphabetical)
- You can even define missing levels, as shown in the next slide


## Factors with ordinal scale

```{r class_sizes_1, fig.width=4.5, fig.height=3.5}
classSizes <- factor(c("big", "small", "huge", "huge", 
    "small","big","small","big"),
    levels = c("small", "normal", "big", "huge"),
    ordered = TRUE)
plot(classSizes)
```


## Calculations with factors in Ordinal scale

- When you have an ordered factor, you can do some calulations with it

```{r class_sizes_2}
classSizes < "big"
sum(classSizes == "huge")
```


## Convert existing factors

When you already have an unorderd factor, you can make it ordered by using the function `ordered()` together with the levels vector

```{r class_sizes_3}
classSizes <- factor(c("big", "small", "huge", "huge",
    "small", "big", "small", "big"))
classSizes <- ordered(classSizes,
                    levels = c("small", "big", "huge"))
classSizes
```


## Working with factors

Factors are used all the time e.g. for defining treated/untreated. That's why R knows how to deal with them so well:

```{r factor_plotting, fig.height = 4, fig.width = 7}
with(ChickWeight, plot(weight ~ Diet))
```


# Lists


## Lists

- A list is an ordered collection of vectors
- These vectors can have **differing types** and **differing lengths**
- Accessing list elements is done with double brackets: `[[]]` or the dollar sign `$` if the elements are named

## List action

```{r list_action}
x <- c(2, 3, 1)
y <- c("foo", "bar")
l <- list(x, y); l
l[[2]]
l[[1]][2]
```


## Named list elements (1)
List can also have named elements

```{r list_named_elements_1}
x <- c(2, 3, 1)
y <- c("foo", "bar")
l <- list("numbers" = x, "words" = y)
l
```


## Named list elements (2)

Accessing named elements can be done in three ways

```{r list_named_elements_2}
l[[2]]        # index
l[["words"]]  # name of element with double brackets
l$words       # name of element with dollar sign
```


## Named list elements (3)

Accessing named elements has its limitations

```{r list_named_elements_3}
select <- "words"
l[[select]] ## OK
l$select ##fails - no element with name "select"
```


## Single versus double brackets on lists

single brackets on a list returns a list; double brackets a vector

```{r list_brackets_1}
l[[2]]
l[2]
l["words"]
```


## Single vs. double brackets on lists (2)

This behaviour can become awkward

```{r list_brackets_2}
l["words"]$words
l[2]["words"][1]$words  ## mind****
```


## Arrays

- Arays are vectors with a dimensions (`dim`) attribute
- Also created using `array()` function
- An array with 2 dimensions is a matrix

```{r arrays1}
x <- 1:10
dim(x) <- c(2, 5)
x
class(x)
```


----

```{r arrays2}
a <- array(data = 1:12, dim = c(2, 3, 2))
# same as "a <- 1:12; dim(a) <- c(2, 3, 2)"
rownames(a) <- c("foo", "bar")
a
class(a)
```


# Dataframes

## data.frame rules all

- In practice you will work with data frames >95% of the time
- Let's meet and greet

```{r data_frame_1, tidy=FALSE}
geneNames <- c("P53","BRCA1","VAMP1", "FHIT")
sig <- c(TRUE, TRUE, FALSE, FALSE)
meanExp <- c(4.5, 7.3, 5.4, 2.4)
genes <- data.frame(
    "name" = geneNames,  
    "significant" = sig,  
    "expression" = meanExp)  
genes
```


-----

```{r data_frame2}
genes[2,1] 		#row 2, element 1
genes[, 1:2] 	#columns 1 and 2
genes[1:2] 		#columns 1 and 2 (!)
```


-----

```{r data_frame3}
genes[1:2,] #row 1 and 2
genes[c("name", "expression")]	#"name" and "expression"
genes$name 	#column "name"
```


## Selections on dataframes summarized

- In general, selections on dataframes are done in this form:
- `my_data[row_sel, col_sel]`
- where `row_sel` and `col_sel` can be 
    - a single index 
    - a numerical vector 
    - a logical vector (of the same length!)
    - empty (for all rows/columns)


## A dataframe is (sort of) a list of vectors

```{r data_frame4}
genes[["name"]] ## select column w. double brackets
class(genes) ## it is NOT a list though
str(genes)
```


# Reading from file

## Loading data frames from file

- In real life, data in dataframes is often loaded from file
- The most used data transfer & storage format is text (tab- or comma-separated)
- Here is an example data set in file ("whale_selenium.txt")

```
whale liver.Se tooth.Se  
1 6.23 140.16  
2 6.79 133.32  
3 7.92 135.34  
...  
19 41.23 206.30  
20 45.47 141.31  
```

## Reading the whale data

```{r echo=FALSE, results="hide"}
setwd("/Users/michiel/Dropbox/courses/ILST_minor_Thema11_intro_R/public/presentations/")
```

```{r data_frame_io1}
whale.selenium <- read.table("data/whale_selenium.txt")
head(whale.selenium)
```

- When loading the data in the standard way, 
    - there is no special consideration for the header line
    - the separator is assumed to be a space
    - the decimal is assumed to be a dot "."

----

- Here, it is specified that
    - the first line is a header line
    - the first colum contains the row names
    
```{r data_frame_io2}
whale.selenium <- read.table(
    file = "data/whale_selenium.txt",
    header = TRUE,
    row.names = 1)
summary(whale.selenium)
```

## What is in that dataframe?

```{r data_frame_sel}
head(whale.selenium, n=3) #have a peek
mean(whale.selenium$liver.Se) #look at a single column
str(whale.selenium) #what is the structure
```


## Ready to rumble

```{r data_frame_io3, fig.width=7, fig.height=4}
plot(
    whale.selenium$liver.Se, whale.selenium$tooth.Se,
    xlab = "liver Selenium", ylab = "tooth Selenium")
abline(lm(whale.selenium$tooth.Se ~
              whale.selenium$liver.Se))
```

----

or, with a smoother:

```{r data_frame_io4, fig.width=7, fig.height=4}
scatter.smooth(
    whale.selenium$liver.Se, whale.selenium$tooth.Se,
    xlab = "liver Selenium", ylab = "tooth Selenium")
abline(lm(whale.selenium$tooth.Se ~ 
              whale.selenium$liver.Se))
```


## Advanced file reading

More advanced file reading will be dealt with in a later presentation.


# Basic DF manipulations


## Changing column names

```{r rename_columns}
names(whale.selenium) <- c("liver", "tooth")
head(whale.selenium, n=2)
##or
colnames(whale.selenium) <- c("brrrr", "gross")
head(whale.selenium, n=2)
```

```{r echo = FALSE, results = "hide"}
colnames(whale.selenium) <- c("liver", "tooth")
```


## Adding columns

You can add columns to an exisiting dataframe

```{r add_columns}
## add simulated stomach data
whale.selenium$stomach <- rnorm(nrow(whale.selenium), 42, 6) 
head(whale.selenium, n=2)
# or
cbind(whale.selenium,
      "a_code" = rep(1:2, nrow(whale.selenium)))
```


## Adding rows: `rbind()`

Adding rows is similar (continued on next slide)

```{r add_rows1}
myData1 <- data.frame(colA = 1:3, colB = c("a", "b", "c"))
myData2 <- data.frame(colA = 4:5, colB = c("d", "e"))
```


----

```{r add_rows2}
myDataComplete <- rbind(myData1, myData2)
myDataComplete
```

Note that the column names of both dataframes need to match for this operation to succeed!


## Getting a summary

```{r summary_demo}
summary(whale.selenium) ## a 6-number summary of each column
```


## Getting the dimensions of a dataframe

```{r dim_demo}
dim(whale.selenium)
```


## A more readable selection

- You can also use `subset()` to make both **column** and **row selections**
- This is a more readable alternative to `[ , ]`
- Note that you don't even need to use quotes

```{r}
##select rows for which Solar.R is available
head(subset(airquality, subset = !is.na(Solar.R)))
```

## `subset()` cont.

```{r}
## select two columns only
head(subset(airquality, select = c(Ozone, Solar.R)))
```

## `subset()` cont.

```{r}
## combine row and colum selection
head(subset(airquality, 
            subset = !is.na(Solar.R), 
            select = c(Ozone, Solar.R)))
```

## `subset()` cont.

```{r}
## shorthand
subset(airquality, Day == 1, select = -Temp)
```

subset() can be used more sophisticated; just GIYF


