---
title: "Dataframe manipulations"
author: "Michiel Noback"
date: "19 November 2015"
output:
  beamer_presentation:
    colortheme: dolphin
    fig_caption: no
    toc: yes
  ioslides_presentation: default
subtitle: Reading, processing and analysing dataframes
job: lecturer at Hanze University of Applied Science
---


```{r setup, cache = F, echo = F, message = F, warning = F}
#comment out for pdf output
source("rmarkdown_knitr_header.R")
```

```{r echo = F}
par(mar=c(2,2,0,0))
```

## Intro

Dataframes are ubiquitous in data analyses using R. There are many functions tailored for DF manipulations; you have already seen `cbind()` and `rbind()`.
In this presentation, we'll explore a few new functions and techniques for working with DFs.

- `with()`
- `subset()`
- `apply()` and its relatives
- `aggregate()`
- `split()`


## `with()` 

- Evaluate an expression in a data environment

```{r with_demo_1, fig.height=3, fig.width=5, fig.show = "hold"}
## this reads really well
with(airquality, boxplot(Ozone ~ Month))
## as opposed to 
#boxplot(airquality$Ozone ~ airquality$Month)
```

-----

- You can also combine expressions within a block of curly braces.

```{r with_demo_2, fig.height=3, fig.width=5}
with(
    airquality, {
        month <- factor(month.abb[Month])
        boxplot(Ozone ~ month)})
```


## `subset()`

Return subsets of vectors, matrices or data frames which meet conditions.

```{r subset_demo_1}
head(
    ## same as airquality[ airquality$Temp >= 80, ]
    subset(airquality, Temp >= 80),
    n = 4
)
```


-----

```{r subset_demo_2}
head(
    subset(airquality, Temp >= 80, select = c(Ozone, Temp)),
    n = 4
)
```

-----


```{r subset_demo_3}
head(
    subset(airquality, is.na(Ozone), select = c(Ozone : Wind)),
    n = 4
)
```



# `apply()` and its relatives

## The apply family

- When you want to do something with 
    - all rows or all columns of a dataframe
    - all values in a vector
    - all elements in a list
- Looping with `for` is very tempting, but often highly inefficient


## Overview

- `apply`: Apply a function over the margins of an array
- `lapply`: Loop over a list and evaluate a function on each element; returns a list of the same length
- `sapply`: Same as lapply but try to simplify the result
- `tapply`: Apply a function over subsets of a vector
- There are more but these are the important ones


## `apply()`

- apply $mean$ to all columns of the built-in `cars` dataset
- apply needs to know 
    1. what DF to apply to
    2. over which margin (columns or rows) 
    3. what function to apply

```{r apply_demo}
apply(cars, 2, mean) # apply over columns
```


## `apply()`: calculate the BMI

- BMI is calculated as $(weight / height ^ 2) * 703$ where weight is in pounds and height in inches.

```{r apply_demo2}
head(women, n=3)
women$bmi <- apply(women, 1, function(x) (x[2] / x[1]^2) * 703)
head(women, n=3)
```


----

- It is not considered good practice to use inline (anonymous) functions because
    - Inline functions make your code less readable
    - Inline functions can not be re-used
- Here is the previous example refactored
    
```{r apply_demo3}
bmi <- function(heightWeight) {
    (heightWeight[2] / heightWeight[1]^2) * 703
}
women$bmi <- apply(women, 1, bmi)
head(women, n=4)
```


-----

- or, maybe better because more generic, but without apply.

```{r apply_demo4}
bmi <- function(height, weight) {
    (weight / height^2) * 703
}
women$bmi2 <- bmi(women$height, women$weight)
head(women, n=4)
```

## passing arguments to the applied function

- Sometimes, the applied function needs to have other arguments passed
- The `...` argument to apply makes this possible (type ?apply)

```{r, results = "hold"}
#sum and power up
spwr <- function(x, p = 2) {sum(x)^p}
df <- data.frame(a = 1:5, b = 6:10)
df
```


-----


```{r, results = "hold"}
apply(df, 1, spwr) # spwr will use default value for p (p = 2)
apply(df, 1, spwr, p = 3) #pass power p = 3 to function spwr
```

## `lapply()`: apply to a list

- `lapply()` applies a function to all elements of a list and returns a list with the same length, each element the result of applying the function

```{r lapply_demo1}
myNumbers = list(
    one = c(1, 3, 4), 
    two = c(3, 2, 6, 1), 
    three = c(5, 7, 6, 8, 9))
lapply(myNumbers, mean)
```


----

- Same, but with `sqrt()` applied
- Note how the nature of the applied function influences the way it is treated

```{r lapply_demo2}
lapply(myNumbers, sqrt)
```


## `sapply()`: apply to a list and try to simplify

- When using the same example as above, but with `sapply`, you get a vector returned
- Note that the elements of the vector do have names attached

```{r sapply_demo}
myNumbers = list(
    one = c(1, 3, 4),
    two = c(3, 2, 6, 1),
    three = c(5, 7, 6, 8, 9))
sapply(myNumbers, mean)
```


## `tapply()`: split and apply

- `tapply()`: Apply a function over subsets of a vector
- in human language: split a vector into groups according to a the levels in a second vector and apply the given function to each group

```{r tapply_demo}
tapply(chickwts$weight, chickwts$feed, mean)
```


## `split()`: split into groups

- Use `split()` when a dataframe needs to be divided depending on the value of some grouping variable. The result is a list, with a member for each grouping value
- Here we have the response of Treated (T) and Untreated (UT) subjects

```{r split_demo1}
myData <- data.frame(
    response = c(5, 8, 4, 5, 9, 3, 6, 7, 3, 6, 5, 2),
    treatment = factor(
        c("UT", "T", "UT", "UT", "T", "UT", "T", "T", "UT", "T", "T", "UT")))
splData <- split(myData, myData$treatment)
str(splData)
```

----

```{r split_demo2, results = 'hold'}
## this trivial example could also have been done with
## boxplot(myData$response ~ myData$treatment)
boxplot(splData$T$response, splData$UT$response, 
        names = c("Treated", "Untreated"))
```

## `aggregate()`: Compute summary statistics of subsets

Splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.

```{r}
aggregate(Temp ~ Month, data = airquality, mean)
```


## many roads lead to Rome

The statements on the next three slides are all essentially the same

```{r}
aggregate(chickwts$weight, by = list(chickwts$feed), FUN = mean)
aggregate(weight ~ feed, data = chickwts, FUN = mean)
```

-----

```{r message=FALSE}
tapply(chickwts$weight, chickwts$feed, mean)
with(chickwts, tapply(weight, feed, mean))
```

-----

```{r message=FALSE}
sapply(split(chickwts, chickwts$feed), function(x){mean(x$weight)})
library(dplyr)
group_by(chickwts, feed) %>% summarise(m = mean(weight))
```

## `merge()` to bring data from two dataframes together

- In many cases, data is distributed over multiple sources (often files).
- The `merge()` function helps you combining these datasets on common identifiers.
- suppose you want to analyse gene class in relation to expression levels
- Note that if two columns have the same name, merge() uses these by default!

----  

```{r}
gene.classes <- data.frame(
    geneID = c("gi:267", "gi:235", "gi:332"),
    class = c("regulator", "metabolism", "structural"))
top.expressed.genes <- data.frame(
    tissue = c("connective", "muscle", "nervous", "epithelial"), 
    gene = c("gi:235", "gi:267", "gi:235", "gi:332"))
merge(top.expressed.genes, gene.classes, by.x="gene", by.y="geneID")

```